---
title: R visualisations using ggplot2 – intermediate
output: github_document 
---

`last revision: 2018-06-29`

> This document is written as an .Rmd document, and then knitted into a GitHub-flavoured markdown document.  
> The source code is available at: https://gitlab.com/stragu/CDS/blob/master/R/ggplot2_intermediate/ggplot2_intermediate.Rmd  
> The published document is available at: https://gitlab.com/stragu/CDS/blob/master/R/ggplot2_intermediate/ggplot2_intermediate.md

If you need to review the installation instructions: https://gitlab.com/stragu/CDS/blob/master/R/Installation.md

## Keep in mind

* Case sensitive
* No spaces in names
* You can use <kbd>Ctrl</kbd>+<kbd>Shift</kbd> to space out your code and <kbd>Ctrl</kbd>+<kbd>Enter</kbd> to run a command.

## Open RStudio

* On CDS computers:
  * Log in with your UQ username and password
  * Make sure you have a working internet connection
  * Go to search the magnifying glass (bottom left)
  * Open the ZENworks application
  * Look for the letter R 
  * Double click on RStudio which will install both R and RStudio 
  * Make sure you have ggplot2 installed

* If you are using your own laptop please open RStudio
  * Make sure you have a working internet connection
  * Make sure you have ggplot2 installed

## Disclaimer

We will assume you are an R intermediate user and that you have used ggplot2 before.

## What are we going to learn?

During this hands-on session, you will:

* install a tool for picking colours;
* customise scales and ranges;
* divide a visualisation in facets;
* explore new geometries;
* modify statistical transformations;
* adjust a geometry's position;
* further modify themes.

## Material

### Setting up

*Exercise 1 - New RStudio Project*

* Click the "Create a project" button (top left cube icon)
* Click "New Directory"
* Click "New Project" ("Empty project" if you have an older version of RStudio)
* In "Directory name", type the name of your project, e.g. "ggplot2_inter"
* Select the folder where to locate your project: the `Documents/RProjects` folder, which you can create if it doesn't exist yet
* Click the "Create Project" button
* Create two folders: "scripts" and "plots".

```{r eval=FALSE}
dir.create("scripts")
dir.create("plots")
```

*Exercise 2 - Setting up: packages and scripts*

* Install and load the package ggplot2
  * Install only if you haven't yet done so using 
        `install.packages("ggplot2")`
  * Load the package using:

```{r}
library(ggplot2)
```

* In the "scripts" folder create (and open) a new R script file called "gapminder_example.R":

```{r eval=FALSE}
file.create("scripts/gapminder_example.R")  # create
file.edit("scripts/gapminder_example.R")    # open
```

Don't forget to add some comments to your header:

```
# Description : create useful plots using ggplot2
# Author: <your name>
# Date: <today's date>
```

### Import data

*Exercise 3 – Read the online data into an R object*

Using the following syntax, how can you read the online CSV data into an R object?

```
gapminder <- ...
```

You have to use the `read.csv()` function, which can take a URL:

```{r}
gapminder <- read.csv(
  file = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
```

*Exercise 4 - Explore the dataset*

You can use `dim()`, `str()`, `head()`, `tail()` and `View()` to explore your dataset.

```{r eval=FALSE}
dim(gapminder)  # what size is the data frame?
str(gapminder)  # how is it structured?
head(gapminder) # have a look at the first few observations
tail(gapminder) # have a look at the last few observations
View(gapminder) # view as a separate spreadsheet
```

Familiarise yourself with the variable names:

```{r}
colnames(gapminder)
```

### Explore data visually

Let's start with a question: Does population grow over the years?

We can make a simple plot with the basics of ggplot2:

```{r}
ggplot(data = gapminder,
       mapping = aes(x = year,
                     y = pop)) +
  geom_point()
```

Remember that only the *data*, the *geometry* and the *aesthetic mappings* always need to be specified.

### Colour, size, shape and other aesthetic attributes

To change the *shape* based on a variable, use `shape = <discrete variable>` inside the `aes()` call.
    
Similarly, to change the *colour* based on a variable, use `colour = <variable>` and `fill = <variable>` inside the `aes()` call.

Finally, if you want to change the size of the geometric object, you can use the `size = <variable>` argument.

*Exercise 5 – Choosing colours*

The `colourpicker` package is a handy addin for RStudio. Install it and use the new "Addins > Colour Picker" menu to pick a different colour for your geometry.

For example, to colour all the points in my geometry (regardless of data):

```{r}
# install the package if you don't have it already:
# install.packages("colourpicker")
ggplot(data = gapminder,
       mapping = aes(x = year,
                     y = pop)) +
  geom_point(colour = "chartreuse3") # change the default colour
```

*Exercise 6 – Colour according to continent*

Modify your plot to colour the points according to the continent variable.

```{r}
ggplot(data = gapminder,
       mapping = aes(x = year,
                     y = pop,
                     colour = continent)) +
  geom_point()
```

### Scales and limits

We could further modify our plot to use a different y axis scale to space out the data:

```{r}
ggplot(data = gapminder,
       mapping = aes(x = year,
                     y = pop,
                     colour = continent)) +
  geom_point() +
  scale_y_log10()
```

> Being explicit about the *argument names* is useful when learning the ins and outs of a function, but as you get more familiar with ggplot2, you can do away with the obvious ones, like `data =` and `mapping =`

To reuse the constant base of our plot (the `ggplot()` call and the point geometry), we can create an object:

```{r}
gg_scatter <- ggplot(gapminder,
                     aes(x = year,
                         y = pop,
                         colour = continent)) +
  geom_point()
```

*Exercise 7 - Modify the axis breaks*

Our x axis is not detailed enough.

We can create a list of all the years when data was collected with the `unique()` function, and modify our plot to add more breaks to our x axis with the `scale_x_continuous()` function. How would you use them?

```{r}
# list all the unique values for year
unique_years <- unique(gapminder$year)
# modify the x axis scale
gg_scatter +
  scale_y_log10() +
  scale_x_continuous(breaks = unique_years)
```

We can further customise a scale with breaks and labels:

```{r}
gg_scatter +
  scale_x_continuous(breaks = unique_years) +
  scale_y_continuous(breaks = c(0, 100000000, 200000000, 500000000, 1000000000),
                     labels = c(0, "100 m", "200 m", "500 m", "1 b"))
```

*Exercise 8 - Modify your y scale range*

To zoom in, we might want to restrict the range by using `ylim()`.

```{r}
gg_scatter +
  ylim(c(0, 360000000))
```
      
Let's talk about that warning message. We could use the `subset()` function to avoid that.

### Histograms

*Exercise 9 – Create a histogram of life expectancy*

Search for the histogram geometry function, and plot the life expectancy. How can we modify the bars?

```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram() # by default, bins = 30
```

Let's change the bin width:

```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram(binwidth = 15)
```

We can also change the number of bins:

```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram(bins = 10)
```

And colour by continent:

```{r}
ggplot(gapminder, aes(x = lifeExp, fill = continent)) +
  geom_histogram(bins = 10)
```

The histogram geometry uses the "stack" **position** by default. It might be more readable if we change it so it uses ratios instead, using `position = "fill"` argument:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 10,
                 position = "fill")
```

We can also make the bars "dodge"" each other:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 10,
                 position = "dodge")
```

### Faceting

An even more readable representation could use **facetting**:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ continent) 
```

### Theming

The legend is probably superfluous. We want to keep the colours, but we use the `theme()` function to customise the look of our plot:
 
```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ continent) +
  theme(legend.position = "none")
```

A pre-built theme function can overwrite some defaults:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ continent) +
  theme_minimal() + # before customising the legend
  theme(legend.position = "none")
```

Try other pre-built themes, like `theme_bw()` and `theme_light()`. You can also start with an empty theme thanks to `theme_void()`, and even install extras with the package `ggthemes`.

### Labels

We can make our plot better with descriptive **labels**:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ continent) +
  theme_minimal() + # before customising the legend
  theme(legend.position = "none") +
  xlab("life expectancy") +
  ylab("count")
```

### Customise a scatterplot

New research question: *How does life expectancy relate to GDP per capita?*

```{r}
ggplot(gapminder,
       aes(x = gdpPercap,
           y = lifeExp)) +
  geom_point() +
  labs(x = "GDP per capita",
       y = "Life expectancy",
       title = "How does GDP relate to life expectancy?") +
  theme_bw()
```

The `labs()` function allows you to group all the labels in one call.

Now, let's make it better: spread the data with a modified **scale**, add a **linear regression**, **colour** the continents and add some **transparency**.

```{r}
ggplot(gapminder,
       aes(x = gdpPercap,
           y = lifeExp)) +
  geom_point(aes(colour = continent),
             alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  labs(x = "GDP per capita",
       y = "Life expectancy",
       title = "How does GDP relate to life expectancy?") +
  theme_bw()
```

### Save your plots

To Save your plots in different formats and resolutions, use 

```{r eval=FALSE}
ggsave("myplot.png", width = 7, height = 5)
```

`ggsave()` allows you to specify more parameters than RStudio's "Export" menu, and is very helpful to export visualisations that will be printed on a poster (`width` and `height` also resize the font, and `DPI` will allow a bigger resolution).

### Other geometries

To count categorical data, use `geom_bar()`:

```{r}
ggplot(gapminder, aes(x = continent)) +
  geom_bar()
```

A simple boxplot to visualise distribution in categories:

```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```

Violin plots might be better at visualising densities:

```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_violin()
```

If the categories get too cramped on the axis, you can rotate them:

```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_violin() +
  theme(axis.text.x = element_text(angle = 90))
```

### Another example

This extra example gives a an idea of how a complex `ggplot2` visualisation might be constructed:

```{r}
ggplot(diamonds,
       aes(x = price,
           y = carat)) +
  geom_point(aes(colour = color),
             alpha = 0.5,
             size = 0.5) +
  scale_color_brewer(palette = "Spectral") +
  geom_smooth(se = FALSE,
              linetype = "dashed",
              colour = "black",
              size = 0.5) +
  facet_wrap(~cut) +
  theme_minimal() +
  labs(x = "price (USD)") +
  coord_flip()
```

### Play time!

* Create a a boxplot of continent vs population
     * Colour and fill by continent
     * try to limit the y axis to see the boxes better
     * let's see if you are able to move the legend to the bottom 

### Modify your plots

Have a look at this [ggplot2 cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf) and experiment customising your plots.

### Close RProject

Closing RStudio will ask you if you want to save your workspace and scripts.

## Important links

* For ggplot2:
    * ggplot2 cheatsheet: https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf
    * Official ggplot2 documentation: http://docs.ggplot2.org/current/
    * Official ggplot2 website: http://ggplot2.tidyverse.org/
    * Chapter on data visualisation in the book _R for Data Science_: http://r4ds.had.co.nz/data-visualisation.html
    * Coding Club's data visualisation tutorial: https://ourcodingclub.github.io/2017/01/29/datavis.html
    * Cookbook for R graphs: http://www.cookbook-r.com/Graphs/
    * STHDA's ggplot2 essentials: http://www.sthda.com/english/wiki/ggplot2-essentials
    * R colours http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
    * Book: Hadley Wickham. ggplot2 Elegant Graphics for Data Analysis Second  Edition. 2016 
https://link-springer-com.ezproxy.library.uq.edu.au/content/pdf/10.1007%2F978-3-319-24277-4.pdf
* R and RStudio in general:
    * RStudio Cheatsheet https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf
    * RStudio online learning https://www.rstudio.com/online-learning/
    * Basic and advanced manuals https://cran.r-project.org/manuals.html
    * Ask about any function or package http://www.rdocumentation.org/
    * If you are looking how-to's or how to fix an error http://stackoverflow.com/questions/tagged/r
    * Lynda.com R training and tutorials https://www.lynda.com/R-training-tutorials/1570-0.html remember to sign in from UQ, https://web.library.uq.edu.au/library-services/training/lyndacom-online-courses
    * ANOVA in R http://homepages.inf.ed.ac.uk/bwebb/statistics/ANOVA_in_R.pdf or https://rcompanion.org/rcompanion/e_01.html
    * If you need an R and/or RStudio workshop/session, please contact the Centre for Digital Scholarship staff to organise one for you. https://web.library.uq.edu.au/locations-hours/centre-digital-scholarship
    * If you have further questions about R, please contact the CDS: cds@library.uq.edu.au
    * Ask questions to other researchers during the weekly Hacky Hour: [https://rcc.uq.edu.au/meetups](https://rcc.uq.edu.au/meetups)
    * Contact your unit's statistician (if you have one?)
