---
title: "R data visualisation with RStudio and ggplot2: intermediate"
output: github_document
date: 2018-09-28
---

> This document is written as an .Rmd document, and then knitted into a GitHub-flavoured markdown document. 
> The source code is available at: https://gitlab.com/stragu/CDS/blob/master/R/ggplot2_intermediate/ggplot2_intermediate.Rmd  
> The published document is available at: https://gitlab.com/stragu/CDS/blob/master/R/ggplot2_intermediate/ggplot2_intermediate.md

If you need to review the installation instructions: https://gitlab.com/stragu/CDS/blob/master/R/Installation.md

## Keep in mind

* Case sensitive
* No spaces in names
* You can use <kbd>Ctrl</kbd>+<kbd>Shift</kbd> to space out your code and <kbd>Ctrl</kbd>+<kbd>Enter</kbd> to run a command.

## Open RStudio

On library computers:

* Log in with your UQ username and password
* Make sure you have a working internet connection
* Go to search the magnifying glass (bottom left)
* Open the ZENworks application
* Look for the letter R 
* Double click on RStudio which will install both R and RStudio 

If you are using your own laptop:

* Make sure you have a working internet connection
* Open RStudio

## Disclaimer

We will assume you are an R intermediate user and that you have used `ggplot2` before.

## What are we going to learn?

During this hands-on session, you will:

* install a tool for picking colours
* customise scales and ranges
* divide a visualisation in facets
* explore new geometries
* modify statistical transformations
* adjust a geometry's position
* further modify themes

## Material

### Setting up

*Exercise 1 - New RStudio Project*

* Click the "Create a project" button (top left cube icon)
* Click "New Directory"
* Click "New Project" ("Empty project" if you have an older version of RStudio)
* In "Directory name", type the name of your project, e.g. "ggplot2_inter"
* Select the folder where to locate your project: the `Documents/RProjects` folder, which you can create if it doesn't exist yet
* Click the "Create Project" button
* Create two folders: "scripts" and "plots".

```{r eval=FALSE}
dir.create("scripts")
dir.create("plots")
```

*Exercise 2 - Setting up: packages and scripts*

* Install and load the package `ggplot2`
  + Install only if you haven't yet done so using: `install.packages("ggplot2")`
  + Load the package using:

```{r}
library(ggplot2)
```

* In the "scripts" folder create (and open) a new R script file called "gapminder_example.R":

```{r eval=FALSE}
file.create("scripts/gapminder_example.R")  # create
file.edit("scripts/gapminder_example.R")    # open
```

Don't forget to add some comments to your header:

```
# Description : ggplot2 intermediate with gapminder data
# Author: <your name>
# Date: <today's date>
```

### Import data

*Exercise 3 – Read the online data into an R object*

Using the following syntax, how can you read the online CSV data into an R object?

```
gapminder <- ...
```

You have to use the `read.csv()` function, which can take a URL:

```{r}
gapminder <- read.csv(
  file = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
```

*Exercise 4 - Explore the dataset*

You can use `dim()`, `str()`, `head()`, `tail()`, `View()` and `summary()` to explore your dataset.

```{r eval=FALSE}
dim(gapminder)     # what size is the data frame?
str(gapminder)     # how is it structured?
head(gapminder)    # have a look at the first few observations
tail(gapminder)    # have a look at the last few observations
View(gapminder)    # view as a separate spreadsheet
summary(gapminder) # summary statistics for each variable
```

Familiarise yourself with the variable names:

```{r}
colnames(gapminder)
```

### Explore data visually

Let's start with a question: Does population grow over the years?

We can make a simple plot with the basics of `ggplot2`:

```{r}
ggplot(data = gapminder,
       mapping = aes(x = year,
                     y = pop)) +
  geom_point()
```

Remember that only the *data*, the *geometry* and the *aesthetic mappings* always need to be specified.

### Aesthetics available

To change the *shape* based on a variable, use `shape = <discrete variable>` inside the `aes()` call.
    
Similarly, to change the *colour* based on a variable, use `colour = <variable>` and `fill = <variable>` inside the `aes()` call.

Finally, if you want to change the size of the geometric object, you can use the `size = <variable>` argument.

#### More control over colours

*Exercise 6 – Colour according to continent*

Modify your plot to colour the points according to the continent variable.

```{r}
ggplot(data = gapminder,
       mapping = aes(x = year,
                     y = pop,
                     colour = continent)) +
  geom_point()
```

This plot uses the default discrete palette.

We can also use different palettes. You can see the ones available in `ggplot2` by looking at the help page of, for example, `scale_colour_brewer()`, under the header "Palettes".

Those are the ColorBrewer palettes, which can be explored online: http://colorbrewer2.org/

If you want a visual overview of the Brewer palettes inside RStudio, install the `RColorBrewer` package and try the following functions:

```{r}
library(RColorBrewer)
display.brewer.all() # all palettes
display.brewer.all(colorblindFriendly = TRUE) # only colourblind-friendly palettes
```

Now, try a different palette for your continents:

```{r}
ggplot(data = gapminder,
       mapping = aes(x = year,
                     y = pop,
                     colour = continent)) +
  geom_point() +
  scale_colour_brewer(palette = "Set2")
```

> **Saving some typing**: To reuse the constant base of our plot (the `ggplot()` call and the point geometry), we can create an object:

```{r}
p <- ggplot(data = gapminder,
            mapping = aes(x = year,
                          y = pop,
                          colour = continent)) +
  geom_point()
```

To use a custom palette, we can use the function `scale_colour_manual()` and provide a list of colour names.

```{r}
p +
  scale_colour_manual(values = c("blue", "red", "purple", "green", "orange"))
```

You can find out about R colour names with the function `colours()`, which prints out a list of their names, but know that you can also use HEX values.

To pick colours more comfortably, we can use the `colourpicker` package, which adds a handy addin in RStudio. Install it and use the new "Addins > Colour Picker" window to create a vector of colours for your custom palette.

### Scale modifiers

We could further modify our plot to use a different y axis scale to space out the data:

```{r}
p +
  scale_y_log10()
```

*Exercise 7 - Modify the axis breaks*

Our x axis is not detailed enough.

We can create a list of all the years when data was collected with the `unique()` function, and modify our plot to add more breaks to our x axis with the `scale_x_continuous()` function. How would you use them?

```{r}
# list all the unique values for year
unique_years <- unique(gapminder$year)
# modify the x axis scale
p +
  scale_x_continuous(breaks = unique_years)
```

We can further customise a scale with breaks and labels:

```{r}
p +
  scale_x_continuous(breaks = unique_years) +
  scale_y_continuous(breaks = c(0, 100000000, 200000000, 500000000, 1000000000),
                     labels = c(0, "100 m", "200 m", "500 m", "1 b"))
```

*Exercise 8 - Modify your y scale range*

To zoom in, we might want to change our axis limits by using `ylim()`.

```{r}
p +
  ylim(c(0, 360000000))
```

Let's talk about that warning message. We could use the `subset()` function to avoid that.

> **Saving some typing:** Being explicit about the **argument names** is useful when learning the ins and outs of a function, but as you get more familiar with ggplot2, you can do away with the obvious ones, like `data =` and `mapping =` (as long as they are used in the right order!).

### Histograms

*Exercise 9 – Create a histogram of life expectancy*

Search for the histogram geometry function, and plot the life expectancy. How can we modify the bars?

```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram() # by default, bins = 30
```

Let's change the bin width:

```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram(binwidth = 15)
```

We can also change the number of bins:

```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram(bins = 10)
```

And colour by continent:

```{r}
ggplot(gapminder, aes(x = lifeExp, fill = continent)) +
  geom_histogram(bins = 10)
```

The histogram geometry uses the "stack" **position** by default. It might be more readable if we change it so it uses ratios instead, using `position = "fill"` argument:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 10,
                 position = "fill")
```

We can also make the bars "dodge"" each other:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 10,
                 position = "dodge")
```

### Faceting

An even more readable representation could use **facetting**:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ continent) 
```

### Theming

The legend is probably superfluous. We want to keep the colours, but we use the `theme()` function to customise the look of our plot:
 
```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ continent) +
  theme(legend.position = "none")
```

A pre-built theme function can overwrite some defaults:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ continent) +
  theme_minimal() + # before customising the legend
  theme(legend.position = "none")
```

Try other pre-built themes, like `theme_bw()` and `theme_light()`. You can also start with an empty theme thanks to `theme_void()`, and even install extras with the package `ggthemes`.

### Labels

We can make our plot better with descriptive **labels**:

```{r}
ggplot(gapminder,
       aes(x = lifeExp,
           fill = continent)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ continent) +
  theme_minimal() + # before customising the legend
  theme(legend.position = "none") +
  xlab("life expectancy") +
  ylab("count")
```

### Customise a scatterplot

New research question: *How does life expectancy relate to GDP per capita?*

```{r}
ggplot(gapminder,
       aes(x = gdpPercap,
           y = lifeExp)) +
  geom_point() +
  labs(x = "GDP per capita",
       y = "Life expectancy",
       title = "How does GDP relate to life expectancy?") +
  theme_bw()
```

The `labs()` function allows you to group all the labels in one call.

Now, let's make it better: spread the data with a modified **scale**, add a **linear regression**, **colour** the continents and add some **transparency**.

```{r}
ggplot(gapminder,
       aes(x = gdpPercap,
           y = lifeExp)) +
  geom_point(aes(colour = continent),
             alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  labs(x = "GDP per capita",
       y = "Life expectancy",
       title = "How does GDP relate to life expectancy?") +
  theme_bw()
```

### Save your plots

To Save your plots in different formats and resolutions, use 

```{r eval=FALSE}
ggsave("myplot.png", width = 7, height = 5)
```

`ggsave()` allows you to specify more parameters than RStudio's "Export" menu, and is very helpful to export visualisations that will be printed on a poster (`width` and `height` also resize the font, and `DPI` will allow a bigger resolution).

### Other geometries

To count categorical data, use `geom_bar()`:

```{r}
ggplot(gapminder, aes(x = continent)) +
  geom_bar()
```

A simple boxplot to visualise distribution in categories:

```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```

Violin plots might be better at visualising densities:

```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_violin()
```

If the categories get too cramped on the axis, you can rotate them:

```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_violin() +
  theme(axis.text.x = element_text(angle = 90))
```

### Another example

This extra example gives a an idea of how a complex `ggplot2` visualisation might be constructed:

```{r}
ggplot(diamonds,
       aes(x = price,
           y = carat)) +
  geom_point(aes(colour = color),
             alpha = 0.5,
             size = 0.5) +
  scale_color_brewer(palette = "Spectral") +
  geom_smooth(se = FALSE,
              linetype = "dashed",
              colour = "black",
              size = 0.5) +
  facet_wrap(~cut) +
  theme_minimal() +
  labs(x = "price (USD)") +
  coord_flip()
```

### Play time!

* Create a a boxplot for each continent's population data
  + Colour and fill by continent
  + Try to limit the y axis to see the boxes better
  + Let's see if you are able to move the legend to the bottom 

### Modify your plots

Have a look at this [ggplot2 cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf) and experiment customising your plots.

### Close RProject

Closing RStudio will ask you if you want to save your workspace and scripts.

## Important links

* For `ggplot2`:
    + ggplot2 cheatsheet: https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf
    + Official ggplot2 documentation: http://docs.ggplot2.org/current/
    + Official ggplot2 website: http://ggplot2.tidyverse.org/
    + Chapter on data visualisation in the book _R for Data Science_: http://r4ds.had.co.nz/data-visualisation.html
    + Coding Club's data visualisation tutorial: https://ourcodingclub.github.io/2017/01/29/datavis.html
    + Cookbook for R graphs: http://www.cookbook-r.com/Graphs/
    + STHDA's ggplot2 essentials: http://www.sthda.com/english/wiki/ggplot2-essentials
    + R colours http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
    + Hadley Wickham's book _ggplot2: Elegant Graphics for Data Analysis_: https://link-springer-com.ezproxy.library.uq.edu.au/content/pdf/10.1007%2F978-3-319-24277-4.pdf
* More ressources for R in general: https://gitlab.com/stragu/CDS/blob/master/R/usefullinks.md