---
title: "R data manipulation with RStudio and dplyr: an introduction"
author: "Stéphane Guillou"
date: "`r Sys.Date()`"
output: github_document
---

Everything we write today will be saved in your R project. Please remember to save it in your H drive or USB if you are using Library computers.

Useful links, exercises and more information on how to continue your R learning are provided in our community resource in this Etherpad: https://frama.link/DSH-dplyr
During the session, it is the only document you need to have open.

## Keep in mind

* R is case sensitive: it will tell the difference between uppercase and lowercase.
* Respect the naming rules for objects (no spaces, does not start with a number...)

### Help

For any dataset or function doubts that you might have, don't forget the three ways of getting help in RStudio:

1. the shortcut command: `?functionname`
2. the help function: `help(functionname)`
3. the keyboard shortcut: press F1 after writing a function name

## What are we going to learn?

In this hands-on session, you will use RStudio and the `dplyr` package to manipulate your data.

Specifically, you will learn how to **explore, filter, reorganise and process** your data with the following verbs:

* `filter()`: pick observations
* `arrange()`: reorder rows
* `select()`: pick variables
* `mutate()`: create new variables
* `summarise()`: collapse to a single summary
* `group_by()`: change the scope of function

## Open RStudio

* If you are using your own laptop please open RStudio
  * If you want to review the installation instructions: https://gitlab.com/stragu/DSH/blob/master/R/Installation.md
* Make sure you have a working internet connection
* On Library computers (the first time takes about 10 min.): 
    * Log in with your UQ username (student if you have two) and password
    * Make sure you have a working internet connection
    * Go to search at bottom left corner (magnifiying glass) 
    * Open the ZENworks application
    * Look for RStudio
    * Double click on RStudio which will install both R and RStudio

## Setting up

### Install the dplyr package

If you don't have it already, you can install dplyr with the command: `install.packages("dplyr")`

> At home, you can install the whole "[tidyverse](https://www.tidyverse.org/)", a meta-package useful for data science: `install.packages("tidyverse")`

### New project

* Click the "File" menu button (top left corner), then "New Project"
* Click "New Directory"
* Click "New Project" ("Empty project" if you have an older version of RStudio)
* In "Directory name", type the name of your project, e.g. "dplyr_intro"
* Select the folder where to locate your project: the `Documents/RProjects` folder, which you can create if it doesn't exist yet.
* Click the "Create Project" button

### Create a script

We will use a script to write code more comfortably.

* Menu: Top left corner, click the green "plus" symbol, or press the shortcut (for Windows/Linux) Ctrl+Shift+N or (for Mac) Cmd+Shift+N. This will open an "Untitled1" file.
* Go to "File > Save" or press (for Windows/Linux) Ctrl+S or (for Mac) Cmd+S. This will ask where you want to save your file and the name of the new file.
* Call your file "process.R"

We can now load the package to access its functions (remember you can use <kbd>Ctrl</kbd>+<kbd>shift</kbd> to execute a command from the script):

```{r}
library(dplyr)
```

> You only need to install a package once (with `install.packages()`), but you need to reload it every time you start a new R session (with `library()`).

### Introducing our data

#### Challenge 1 – import data

Let's import and explore our data.

1. read the data into an object called "gapminder", using `read.csv()`:

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
```

2. Explore the gapminder dataset using `dim()` and `str()`

How can we get the dataframe's variable names? There are two ways: `names(gapminder)` returns the names regardless of the object type, such as list, vector, data.frame etc., whereas `colnames(gapminder)` returns the variable names for matrix-like objects, such as matrices, dataframes...

To select one specific column in the dataframe, you can use the dollar sign in: `gapminder$year`. For example, try these:

```{r}
nlevels(gapminder$country)
class(gapminder$country)
```

If we want nicer printing, we can convert our dataframe to a "tibble". You can install the `tibble` package with `install.packages(tibble)` and then run:

```{r}
library(tibble)
gapminder <- as_tibble(gapminder)
```

## Basic dplyr verbs

The R package `dplyr` was developed by Hadley Wickham for data manipulation.

The book _[R for Data Science](http://r4ds.had.co.nz)_ introduces the package as follows:

> You are going to learn the five key dplyr functions that allow you to solve the vast majority of your data manipulation challenges:
> 
> * Pick observations by their values with `filter()`.
> * Reorder the rows with `arrange()`.
> * Pick variables by their names with `select()`.
> * Create new variables with functions of existing variables with `mutate()`.
> * Collapse many values down to a single summary with `summarise()`.
> 
> These can all be used in conjunction with `group_by()` which changes the scope of each function from operating on the entire dataset to operating on it group-by-group. These six functions provide the **verbs for a language of data manipulation**.

To use the verbs to their full extent, we will need **pipes** and **logical operators**, which we will introduce as we go.

### Pick observations with `filter()`

The `filter()` function allows use to pick observations depending on one ore several conditions.

**Logical operators** allow us to **check for a condition**. Remember: `=` is to assign a value to a variable, `==` is to check for a condition.

* `==` equal
* `!=` different 
* `>` greater than
* `<` smaller than
* `>=` greater or equal
* `<=` smaller or equal

Filter the observations for Australia, using `filter()` and a logical operator:

```{r}
australia <- filter(gapminder, country == "Australia")
australia
```

Filter the rows that have a life expectancy `lifeExp` greater than 80 years:

```{r}
life80 <- filter(gapminder, lifeExp > 80)
dim(life80)
```

### Reorder rows with `arrange()`

If we want to have a look at the entries with highest GDP per capita:

```{r}
arrange(gapminder, desc(gdpPercap))
```

### Pick variables with `select()`

Select allows use to pick variable (i.e. columns) from the dataset. For example, to only keep the data about year, country and GDP per capita:

```{r}
(gap_small <- select(gapminder, year, country, gdpPercap))
```

We wrap it in parentheses so it also prints to screen.

If we only want this data for 1997, we can associate `select()` to `filter()`:

```{r}
gap_small_97 <- filter(gap_small, year == 1997)
```

We can make our code more readable and avoid creating useless intermediate objects by **piping** commands into each other. The pipe symbol `%>%` **strings commands together**, using the left-side output as the first argument of the right-side function.

This command:

```{r}
summary(gapminder)
```

Is equivalent to:

```{r}
gapminder %>% summary()
```

Here's another example with the `filter()` verb:

```{r}
gapminder %>%
  filter(country != "France")
```

... becomes:

```{r}
filter(gapminder, country != "France")
```

To do what we did previously in one single command, using the pipe:

```{r}
gap_small_97 <- gapminder %>%
    select(year, country, gdpPercap) %>%
    filter(year == 1997)
```

The pipe operator can be read as "then" and makes the code a lot **more readable** than when nesting functions into each other, and avoids the creation of several intermediate objects.

From now on, we'll use this syntax.

#### Challenge 2 – a tiny dataset

Select the 2002 life expectancy observation for Eritrea (and remove the rest of the variables).

```{r}
eritrea_2002 <- gapminder %>%
    select(year, country, lifeExp) %>%
    filter(country == "Eritrea", year == 2002)
```

### Create new variables with `mutate()`

Have a look at what the verb `mutate()` can do with `?mutate`.

Let's see what the two following variables can be used for:

```{r}
gapminder %>%
    select(gdpPercap, pop)
```

#### Challenge 3 – mutate the GDP

Use `mutate()` to create a `gdp` variable.

Name your new dataset `gap_gdp`. When finished, `dim(gap_gdp)` should result in `1704 7`.

Hint: use the `*` operator within `mutate()`.

```{r}
gap_gdp <- gapminder %>%
    mutate(gdp = gdpPercap * pop)
dim(gap_gdp)
head(gap_gdp)
```

Reuse a variable computed by 'mutate()' straight away:

```{r}
(gap_gdp <- gapminder %>%
    mutate(gdp = gdpPercap * pop, gdpMil = gdp / 10^6))
```

### `group_by()` and `summarise()`

`group_by()` changes the scope of each function from operating on the entire dataset to operating on it group-by-group. For example, to group by continents:

```{r}
gap_continents <- gapminder %>%
    group_by(continent)
head(gap_continents)
```

`summarise()` collapses many values down to a single summary. For example, to find the mean life expectancy for the whole dataset:

```{r}
gapminder %>%
  summarise(meanLE = mean(lifeExp))
```

Associating the two functions makes it more interesting. To find out the mean life expectancy for each continent in 2007, we can do the following:

```{r}
gapminder %>% 
    filter(year == 2007) %>%
    group_by(continent) %>%
    summarise(meanLE = mean(lifeExp))
```

#### Challenge 4 – max life expectancy

Group by country, and find out the maximum life expectancy ever recorded

Hint: `?max`

```{r}
gapminder %>% 
    group_by(country) %>%
    summarise(maxLE = max(lifeExp))
```

## More examples

Another example of new variable with `mutate()`:

```{r}
starwars %>% 
  mutate(name, bmi = mass / ((height / 100)  ^ 2)) %>%
  select(name:mass, bmi)
```

And a more complex processing of a dataset:

```{r}
starwars %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mass = mean(mass, na.rm = TRUE)
  ) %>%
  filter(n > 1)
```

An example of data manipulation and data visualisation in the same command:

```{r}
# increase in population in Europe
library(ggplot2)
gapminder %>% 
  filter(continent == "Europe") %>%
  group_by(year) %>% 
  summarise(sum = sum(pop)) %>% 
  ggplot(aes(x = year,
             y = sum)) +
  geom_line()
```

And another one, still using our gapminder dataset:

```{r}
# top and bottom variations in life expectancy
gapminder %>% 
  group_by(country) %>% 
  summarise(maxLifeExp = max(lifeExp),
            minLifeExp = min(lifeExp)) %>% 
  mutate(dif = maxLifeExp - minLifeExp) %>% 
  arrange(desc(dif)) %>% 
  slice(1:10, (nrow(.)-10):nrow(.)) %>% 
  ggplot(aes(x = reorder(country, dif), y = dif)) +
  geom_col() +
  coord_flip()
```

## Close project

If you want to close RStudio, make sure you save your script first.

You can then

## What next?

Look at our compilation of resources: https://gitlab.com/stragu/DSH/blob/master/R/usefullinks.md