---
title: Heatmaps in R for intermediate users
date: 2018-06-19
output:
  github_document: default
---

> This document is edited as an R markdown file, and regularly exported as a GitHub document.
> The source code is [here](https://gitlab.com/stragu/CDS/blob/master/R/heatmaps_intermediate.Rmd)
> The published printer-friendly version is [here](https://gitlab.com/stragu/CDS/blob/master/R/heatmaps/heatmaps_intermediate.md)

If you want to review the installation instructions: https://gitlab.com/stragu/CDS/blob/master/R/Installation.md

Everything we write today will be saved in your R Project. Please remember to save it in your H drive or USB if you use the University computers.

## Keep in mind

* Case sensitive
* No spaces in names
* Use <kbd>Ctrl</kbd>+<kbd>Enter</kbd> to run a command, and <kbd>Ctrl</kbd>+<kbd>Shift</kbd> to make your code more readable.

## Open RStudio

* If you are using your own laptop please open RStudio
  * Make sure you have a working internet connection

* On CDS computers:
  * Log in with your UQ username and password
  * Make sure you have a working internet connection
  * Open the ZENworks application
  * Look for the letter "R"
  * Double click on RStudio which will install both R and RStudio 

## What are we going to learn?

During this session, you will:

* Learn how to produce a simple heatmap with the base package `heatmap()`;
* Learn about alternatives to produce more complex heatmaps, like `heatmap.2()` and `pheatmap()`;
* Learn how to produce a rudimentary heatmap with the `ggplot2` package.

A **heatmap** is a way of visualising a table of numbers, where you substitute the numbers with colored cells. It’s useful for finding highs and lows, and see patterns more clearly.

## Disclaimer

We will assume you are an R intermediate user and that you have used RStudio before.

## Material

### RStudio Project

**Exercise 1 - New RStudio Project**

* New project:
  * Click the "New project" menu icon
  * Click "New Directory"
  * Click "New Project" ("New empty project" if you have an older version of RStudio)
  * In Directory name type the name of your project, e.g. "heatmaps" (Browse and select a folder where to locate your project, in our case the RProjects folder. If you don't have an RProjects folder, create it.)
  * Click the "Create Project" button
* Create new folders with the following commands:

```{r eval=FALSE}
dir.create("scripts")
dir.create("data")
dir.create("plots")
```

**Exercise 2 - setting up**

* create a new R script called "heatmaps.R" in the "scripts" folder

```{r eval=FALSE}
file.create("scripts/heatmaps.R")
file.edit("scripts/heatmaps.R")
```

### Method 1: the base `heatmap()` function

As a first example, we will use a built-in datasets called `mtcars`.

*Step 1: explore the data*

```{r eval=FALSE}
?mtcars
dim(mtcars)
str(mtcars)
head(mtcars)
View(mtcars)
class(mtcars)
```

*Step 2: prepare data*

The data is a dataframe, but it has to be a **numeric matrix** to make your heatmap. Dataframes can contain variables with different data classes

**Exercise 3 – Convert the mtcars data into a matrix**

```{r}
class(mtcars)
mtcars_matrix <- data.matrix(mtcars) # convert a DF to a numeric matrix
class(mtcars_matrix)
```

*Step 3: make a heatmap*

We are going to use the `heatmap()` function.

**Exercise 4 – look for the `heatmap()` help**

Read the description of the `scale` argument in particular.

```{r}
?heatmap
```

Create a first heatmap:

```{r}
heatmap(mtcars_matrix)
```

**Scale** is important: the values should be centered and scaled in either rows or columns. In our case, the variables are contained in columns.

```{r}
heatmap(mtcars_matrix, scale = "column")
```

#### Colours

The default heatmap palette goes from red for low values to white for high values. You can however replace the default palette and use different colours, and different numbers of levels.

For example, in the palette function `cm.colors(n)`, n is the number of levels (>= 1) contained in the cyan-to-magenta palette. It can be used in the `col` argument.

```{r}
heatmap(mtcars_matrix,
        col = cm.colors(15),
        scale = "column",
        margins = c(8, 8))
```

You can try other functions, like `rainbow()` or `terrain.colors()`.

#### Remove dendrograms

Dendrograms don't really make sense for this dataset. `Rowv` and `Colv` can be set to `NA` to remove them.

```{r}
heatmap(mtcars_matrix,
        col = cm.colors(100),
        scale = "column",
        Rowv = NA,
        Colv = NA)
```

#### Clean the environment

We can start with a fresh environment, using:

```{r}
rm(list = ls())
```

### Method 2: `gplots::heatmap.2()`

If you don't have the `gplots` package yet, use `install.packages("gplots")`. We will also need the `RColorBrewer`, so use `install.packages("RColorBrewer")` if you don't have it installed already.

```{r include=FALSE}
library(gplots)
library(RColorBrewer)
```

#### Protein data example

*Step 1: data*

The dataset contains observations for 63 proteins in three control experiments and three experiments where cells are treated with a growth factor.

```{r}
rawdata <- read.csv("https://raw.githubusercontent.com/ab604/heatmap/master/leanne_testdata.csv")
?heatmap.2
```

This `gplots` heatmap function provides a number of extensions to the standard R heatmap function.

*Step 2: explore the data*

```{r eval=FALSE}
dim(rawdata)
str(rawdata)
head(rawdata)
View(rawdata)
```

*Step 3: data munging*

It’s important to note that a lot of visualisations involve gathering and preparing data. Rarely do you get data exactly how you need it, so you should expect to do some data munging before producing the visuals.

```{r}
rawdata$Row_ID <- NULL
rawdata$T.test <- NULL
colnames(rawdata) <- c(paste("Control", 1:3, sep = "_"), 
                       paste("Treatment", 1:3, sep = "_"))
```

*Step 4: convert the dataframe into a matrix*

```{r}
class(rawdata)
data_matrix <- data.matrix(rawdata)
head(data_matrix)
class(data_matrix)
```

*Step 5: more munging*

```{r eval=FALSE}
?scale
```

`scale` is generic function whose default method centers and/or scales the columns of a numeric matrix. We **normalise** the data using `scale()`, **transposing** using `t()` to perform row-wise normalisation of the protein data. This normalises the data across each experiment so we can compare them to each other.
Finally, we transpose the data back to the original form.

```{r}
# Normalise data, transpose for row wise normalisation
data_scaled_t <- scale(t(rawdata))
# check normalisation for each protein by:
  # checking that means are zero
round(colMeans(data_scaled_t), 1)
  # check std devs are 1
apply(data_scaled_t, 2, sd)
# transpose to original form
data_scaled <- t(data_scaled_t)
```

*Step 5: create heatmaps*

First, let's experiment with the base `heatmap()` function.

```{r}
heatmap(data_scaled)
```

```{r}
heatmap(data_scaled,
        col = heat.colors(8), 
        margins = c(8,12)) # , reveal col labels, reduce width
```

Let's create a new palette with 25 increments:

```{r}
my_palette <- colorRampPalette(c("blue",
                                 "white",
                                 "red"))(n = 25)
```

Now, let's use `heatmap.2()`:

```{r}
heatmap.2(data_scaled,                  # normalised data
          density.info = "histogram",   # plot histogram of data and colour key
          trace = "none",               # turn off trace lines from heatmap
          col = my_palette)             # use my colour scheme
```

Fix a few things and add a few extras:

```{r}
par(cex.main = 0.75)                  # shrink title fonts on plot
heatmap.2(data_scaled,
          density.info = "histogram",
          dendrogram = "row",         # only row dendrogram
          trace = "none",
          col = my_palette,
          main = "Difference between proteins",
          margins = c(8, 5),          # more space from border
          offsetRow = 1,              # space between plot and labels
          keysize = 2,                # make key and histogram fit
          cexRow = 0.5,               # amend row font
          cexCol = 0.75)              # amend column font
```

Clean up the environment with:

```{r}
rm(list = ls())
```

#### (optional) Genechip data example

The `BiocInstaller` package allows to use the `biocLite()` function to install extra packages, including the `affy` package which we need in this example (it contains the data).
Install BiocInstaller with `source("https://bioconductor.org/biocLite.R")`, and then `affy` with the command `biocLite("affy")`

```{r eval=FALSE}
library(affy)
data(SpikeIn)
pms <- SpikeIn@pm
str(pms)
class(pms)
# just the data, scaled across rows
par(cex.main = 0.75)  
labs <- colnames(pms)
heatmap.2(pms, 
          col = rev(heat.colors(16)), 
          main = "SpikeIn@pm",
          xlab = "Relative Concentration", 
          ylab="Probeset",
          trace = "none",
          labCol = "",
          add.expr = text( x = seq_along(labs), y = -2, srt = 45,
                           labels = labs, xpd = TRUE),
          scale="row")
```

Clean up your environment with:

```{r}
rm(list = ls())
```

### Method 3: `pheatmap()`

If you don't have it already, install `pheatmap` with `install.packages("pheatmap")`.
Load the packages with:

```{r}
library(grid)     # this is a base package  
library(pheatmap) # stands for "pretty heatmaps"
```

How does `pheatmap` differ?

```{r eval=FALSE}
?pheatmap
```

> A function to draw clustered heatmaps where one has better control over some graphical parameters such as cell size, etc.

Create your data matrix:

```{r}
set.seed(100)
d <- matrix(rnorm(25), 5, 5)
colnames(d) <- paste("Treatment", 1:5, sep = "")
rownames(d) <- paste("Gene", 1:5, sep = "")
```

Try it out:
```{r}
pheatmap(d, 
         main = "Pretty heatmap",
         cellwidth =  50,
         cellheight = 30,
         fontsize = 12)
```

You can save your plot with an extra argument:

```{r eval=FALSE}
pheatmap(d, 
         main = "Pretty heatmap",
         cellwidth =  50,
         cellheight = 30,
         fontsize = 12,
         filename = "plots/heatmap.pdf")
```

Clean up your environment with:

```{r}
rm(list = ls())
```

### Method 4: a dataframe in `ggplot2`

Load the necessary libraries:

```{r}
library(reshape2)
library(ggplot2)
```

Labels the rows and columns:

```{r}
name_genes <- paste("GEN", LETTERS[1:20], sep="_") # rows
name_patients <- paste("PATIENT", 1:20, sep="_")  # columns
```

Generation of dataframe:

```{r}
value_expression <- data.frame(genes = name_genes, 
                               matrix(rnorm(400, 2, 1.8),
                                      nrow = 20,
                                      ncol = 20))
names(value_expression)[2:21] <- name_patients
```

Melt the dataframe:

```{r}
df_heatmap <- melt(value_expression, id.vars = "genes")
names(df_heatmap)[2:3] <- c("patient", "expression_level")
head(df_heatmap)
```

Create a basic heatmaps from the dataframe:

```{r}
ggplot(df_heatmap, aes(patient, genes)) +
  geom_tile(aes(fill = expression_level), color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  ylab("List of genes") +
  xlab("List of patients") +
  theme(legend.title = element_text(size = 10),
        legend.text = element_text(size = 12),
        plot.title = element_text(size=16),
        axis.title=element_text(size=14,face="bold"),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(fill = "Expression level")
```

This ggplot2 method does not allow to create dendrograms.

Clean up your environment with:

```{r}
rm(list = ls())
```

### (optional) Method 5: `ComplexHeatmap::Heatmap`

URL: https://www.bioconductor.org/packages/devel/bioc/vignettes/ComplexHeatmap/inst/doc/s2.single_heatmap.html

*Step 1*

```{r}
# source("https://bioconductor.org/biocLite.R")
# biocLite("ComplexHeatmap") 
library(ComplexHeatmap)
# install.packages("circlize")
library(circlize)
```

Create a data matrix:

```{r}
set.seed(123)
mat <- cbind(rbind(matrix(rnorm(16, -1), 4), matrix(rnorm(32, 1), 8)),
            rbind(matrix(rnorm(24, 1), 4), matrix(rnorm(48, -1), 8)))
```

*Step 2: munging*

Permute the rows and columns
```{r}
mat <- mat[sample(nrow(mat), nrow(mat)), sample(ncol(mat), ncol(mat))]
rownames(mat) <- paste0("R", 1:12)
colnames(mat) <- paste0("C", 1:10)
```

*Step 3: make a heatmap*

```{r}
?Heatmap
Heatmap(mat)
```

```{r}
Heatmap(mat, 
        col = colorRamp2(c(-3, 0, 3), c("brown", "white", "yellow")), 
        cluster_rows = FALSE, 
        cluster_columns = FALSE,
        heatmap_legend_param = list(title = "Values"))
```

```{r}
Heatmap(mat, 
        col = colorRamp2(c(-3, 0, 3), c("brown", "white", "yellow")), 
        heatmap_legend_param = list(title = "Values"))
```

```{r}
mat_with_na <- mat
mat_with_na[sample(c(TRUE, FALSE), nrow(mat)*ncol(mat), replace = TRUE, prob = c(1, 9))] <- NA
Heatmap(mat_with_na, 
        col = topo.colors(100),
        na_col = "orange", 
        clustering_distance_rows = "pearson",
        heatmap_legend_param = list(title = "Values"))
```


```{r}
Heatmap(mat, 
        name = "values", 
        row_names_side = "left", 
        row_dend_side = "right", 
        column_names_side = "top", 
        column_dend_side = "bottom")
```

```{r}
Heatmap(mat, 
        name = "values", 
        row_names_side = "left", 
        row_dend_side = "right", 
        column_names_side = "top", 
        column_dend_side = "bottom",
        km = 2)
```

```{r}
h1 <- Heatmap(mat, 
        name = "values", 
        #col = topo.colors(50),
        color_space = "sRGB",
        row_dend_width = unit(2, "cm"),
        column_dend_height = unit(2, "cm"),
        row_dend_reorder = TRUE,
        column_dend_reorder = TRUE,
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 7),
        column_names_max_height = unit(2, "cm"),
        row_names_max_width = unit(9, "cm"),
        column_title = "This is a complex heatmap")
draw(h1, heatmap_legend_side = "left")
```

Clean my environment with:

```{r}
rm(list = ls())
```

## Close Rproject

When closing RStudio, you should be prompted to save your workspace.

## Important links

* RStudio Cheatsheet https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf
* RStudio online learning https://www.rstudio.com/online-learning/
* Basic and advanced manuals https://cran.r-project.org/manuals.html
* Ask about any function or package http://www.rdocumentation.org/
* If you are looking how-to's or how to fix an error http://stackoverflow.com/questions/tagged/r
* Lynda.com R training and tutorials https://www.lynda.com/R-training-tutorials/1570-0.html remember to sign in with your organisational portal,https://web.library.uq.edu.au/library-services/training/lyndacom-online-courses
* R colours http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
* Book: Hadley Wickham. ggplot2 Elegant Graphics for Data Analysis Second  Edition. 2016 
https://link-springer-com.ezproxy.library.uq.edu.au/content/pdf/10.1007%2F978-3-319-24277-4.pdf
* Examples of heatmaps: https://flowingdata.com/2010/01/21/how-to-make-a-heatmap-a-quick-and-easy-solution/
 https://rpubs.com/ab604/98032  https://stackoverflow.com/questions/15505607/diagonal-labels-orientation-on-x-axis-in-heatmaps  https://www.bioconductor.org/packages/devel/bioc/vignettes/ComplexHeatmap/inst/doc/s2.single_heatmap.html

* If you need an R and/or RStudio workshop/session, please contact Centre for Digital Scholarship staff to organise one for you. https://web.library.uq.edu.au/locations-hours/centre-digital-scholarship
* If you have further questions, please contact me p.martinez at uq.edu.au 
