---
title: 'R with RStudio: getting started'
author: "Stéphane Guillou"
date: "`r Sys.Date()`"
output: github_document
---

> These notes are available on GitLab: https://gitlab.com/stragu/DSH/blob/master/R/rstudio_intro/rstudio_intro.md

## R + RStudio

The [R programming language](https://cran.r-project.org/) is a language used for calculations, statistics, visualisations and many more data science tasks.

[RStudio](https://rstudio.com/products/rstudio/) is an open source Integrated Development Environment (IDE) for R, which means it provides many features on top of R to make it easier to write and run code.

R's main strong points are:

* **Open Source**: you can install it anywhere and adapt it to your needs;
* **Reproducibility**: makes an analysis repeatable by detailing the process in a script;
* **Customisable**: being a programming language, you can create your own custom tools;
* **Big data**: it can handle very large datasets;
* **Large ecosystem**: packages allow you to extend R for thousands of different analyses.

The learning curve will be steeper than point-and-click tools, but as far as programming languages go, R is more user-friendly than others.

## Installation

For this course, you need to have both R and RStudio installed ([installation instructions](https://gitlab.com/stragu/DSH/blob/master/R/Installation.md).

## Open RStudio

* If you are using your own laptop please open RStudio
  * Make sure you have a working Internet connection
* On Library computers:
  * Log in with your UQ username and password (if you are both staff and student, use your student account)
  * Make sure you have a working Internet connection
  * Go to search at bottom left corner (magnifiying glass) 
  * Open the ZENworks application
  * Search for "RStudio"
  * Double-click on RStudio which will install both R and RStudio

## What are we going to learn?

This session is designed to get straight into using R in a short amount of time, which is why we won't spend too much time on the smaller details that make the language.

During this session, you will:

* Create a project for data analysis
* Create a folder structure
* Know where to find help
* Learn about a few useful functions
* Create a script
* Import a dataset
* Understand the different RStudio panels
* Use a few shortcuts
* Know how to extend R with packages
* Generate a data visualisation

## R Projects

Let's first create a new project:

* Click the "File" menu button (top left corner), then "New Project"
* Click "New Directory"
* Click "New Project"
* In "Directory name", type the name of your project, for example "YYYY-MM-DD_rstudio-intro"
* Browse and select a folder where to locate your project (`~` is your home directory). For example, a folder called "r-projects".
* Click the "Create Project" button

> R Projects make your work with R more straight forward, as they allow you to segregate your different projects in separate folders. You can create a .Rproj file in a new directory or an existing directory that already has R code and data. Everything then happens by default in this directory. The .Rproj file stores information about your project options, and allows you to go straight back to your work.

## Maths and objects

The **console** (usually at the bottom right in RStudio) is where most of the action happens. In the console, we can use R interactively. We write a **command** and then **execute** it by pressing <kbd>Enter</kbd>.

In its most basic use, R can be a calculator. Try executing the following commands:

```{r}
10 - 2
3 * 4
2 + 10 / 5
11^6
```

Those symbols are called "binary operators": we can use them to multiply, divide, add, subtract and exponentiate. Once we execute the command (the "input"), we can see the result in the console (the "output").

What if we want to keep reusing the same value? We can store data by creating **objects**, and assigning values to them with the **assignment operator** `<-`:

```{r}
num1 <- 42
num2 <- num1 / 9
num2
```

We can also store text data:

```{r}
sentence <- "Hello World!"
sentence
```

You should now see your objects listed in you **environment pane** (top right).

As you can see, you can store different kinds of data as objects. If you want to store text data (a "string of characters"), you have to use quotes around them.

> You can use the shortcut <kbd>Alt</kbd>+<kbd>-</kbd> to type the assignement operator quicker.

## Functions

An R **function** is a little program that does a particular job. It usually looks like this:

```
<functionname>(<argument(s)>)
```

**Arguments** tell the function what to do. Some functions don't need arguments, others need one or several, but they always need the parentheses after their name.

For example, try running the following command:

```{r}
class(num1)
```

The `class()` function tells us what class of data we are dealing with. Here, we use the object `num1` as the first and only argument.

> If you scroll back to the top of your console, you will now be able to spot functions in the text.

### Help

What if we want to learn more about a function?

There are two main ways to find **help** about a specific function in RStudio:

1. the shortcut command: `?functionname`
1. the keyboard shortcut: press <kbd>F1</kbd> with your cursor in a function name

Let's look through the documentation for the `class()` function:

```r
?class
```

There is quite a lot of information in there, but the most important bits are:

* **Description**: general description of the function(s)
* **Usage**: overview of what syntax can be used
* **Arguments**: description of what each argument is
* **Examples**: some examples that demonstrate what is possible

#### Challenge 1 – Finding help

Use the help pages to find out what these functions do, and try executing commands with them:

1. `c()`
1. `rep.int()`
1. `mean()`
1. `rm()`

`c()` concatenates the arguments into a vector. In other words, it takes any number of arguments (hence the `...`), and stores all those values together. For example, let's store the ages of our pet dogs in a new object:

```{r}
ages <- c(4, 10, 2, NA, 3)
```

We can now reuse this vector, and calculate their human age:

```{r}
ages * 7
```

`rep.int()` also creates vectors, but it is designed to easily replicate values. For example, if you find something very funny:

```{r}
rep.int("Ha!", 30)
```

`mean()` returns the mean of a vector of numbers:

```{r}
mean(ages)
```

What happened there?

We have an NA value in the vector, which means the function can't tell what the mean is. If we want to change this default behaviour, we can use an extra argument: `na.rm`, which stands for "remove NAs".

```{r}
mean(ages, na.rm = TRUE)
```

> We don't need to specify the names of the arguments as long as we use them in order. In our last command, if we hadn't named the `na.rm` argument, R would have understood `TRUE` to be the value for the `trim` argument!

`rm()` removes an object from your environment (`remove()` and `rm()` point to the same function). For example:

```{r}
rm(num1)
```

> R does not check if you are sure you want to remove something! As a programming language, it does what you ask it to do, which means you might have to be more careful. But you'll see later on that, when working with scripts, this is less of a problem.

Let's do some more complex operations by combining two functions:

`ls()` returns a character vector: it contains the names of all the objects in the current environment (i.e. the objects we created in this R session). Is there a way we could combine it with `rm()`?

You can remove *all* the objects in the environment by using `ls()` as the value for the `list` argument:

```{r}
rm(list = ls())
```

We are nesting a function inside another one. More precisely, we are using the output of the `ls()` function as the value passed on to the `list` argument in the `rm()` function.

### More help

We've practised how to find help about functions we know the name of.
What if we don't know what the function is called? Or if we want general help about R?

* The function `help.start()` is a good starting point: it opens a browser of official R help.
* If you want to search for a word in all the documentation, you can use the `??` syntax. For example, try executing `??anova`.
* Finally, you will often go to your web browser and search for a particular question, or a specific error message: most times, there already is an answer somewhere on the Internet. The challenge is to ask the right question!

## Create a folder structure

To keep it tidy, we are creating 3 folders in our project directory:

* scripts
* data
* plots

For that, we use the function `dir.create()`:

```r
dir.create("scripts")
dir.create("data")
dir.create("plots")
```

> You can recall your recent commands with the up arrow, which is especially useful to correct typos or slightly modify a long command.

## Scripts

Scripts are simple text files that contain R code. They are useful for:

* saving a set of commands for later use (and executing it in one click)
* making research reproducible
* making writing and reading code more comfortable
* documenting the code with comments, and
* sharing your work with peers

Let's create a new R script with a command:

```r
file.create("scripts/process.R")
```

> All the file paths are **relative** to our current working directory, i.e. the project directory. To use an **absolute** file path, we can start with `/`.

To edit the new script, use the `file.edit()` function. Try using the <kbd>Tab</kbd> key to autocomplete your function name and your file path!

```r
file.edit("scripts/process.R")
```

This opens our fourth pannel in RStudio: the **source pannel**.

### Many ways to do one thing

As in many programs, there are many ways to achieve one thing.

For example, we used commands to create and edit a script, but we could also:

* use the shortcut <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>N</kbd>
* use the top left drop-down menus

Learning how to use functions rather than the graphical user interface (GUI) will allow you to integrate them in scripts, and will sometimes help you to do things faster.

### Comments

We should start with a couple of **comments**, to document our script. Comments start with `#`, and will be ignored by R:

```r
# Description: Introduction to R and RStudio
# Author: <your name>
# Date: <today's date>
```

### Syntax highlighting

Now, add a command to your script:

```{r}
vect <- c(3, "Hi!")
```

Notice the colours? This is called **syntax highlighting**. This is one of the many ways RStudio makes it more comfortable to work with R. The code is more readable when working in a script.

> While editing your script, you can run the current command (or the selected block of code) by using <kbd>Ctrl</kbd>+<kbd>Enter</kbd>. Remember to save your script regularly with the shortcut <kbd>Ctrl</kbd>+<kbd>S</kbd>. You can find more shortcuts with <kbd>Alt</kbd>+<kbd>Shift</kbd>+<kbd>K</kbd>, or the menu "Tools > Keyboard Shortcuts Help".

## Import data

### Challenge 2 – Import data

Copy and paste the following two commands into your script:

```r
download.file(url = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv",
  destfile = "data/gapminderdata.csv")
gapminder <- read.csv("data/gapminderdata.csv")
```
```{r include=FALSE}
gapminder <- read.csv("https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
```

What do you think they do? Describe each one in detail, and try executing them.

## Explore data

We have downloaded a CSV file from the Internet, and read it into an object called `gapminder`.

You can type the name of your new object to print it to screen:

```{r eval=FALSE}
gapminder
```

That's a lot of lines printed to your console. To have a look at the first few lines only, we can use the `head()` function:

```{r}
head(gapminder)
```

Now let's use a few functions to learn more about our dataset:

```{r}
class(gapminder) # what kind of object is it stored as?
nrow(gapminder) # how many rows?
ncol(gapminder) # how many columns?
dim(gapminder) # rows and columns
names(gapminder) # variable names
```

All the information we just saw (and more) is available with one single function:

```{r}
str(gapminder) # general structure
```

> The RStudio's environment pannel already shows us some of that information (click on the blue arrow next to the object name).

And to explore the data in a viewer, run the following:

```r
View(gapminder) # spreadsheet-like view in new tab
```

This viewer allows you to explore your data by scrolling through, searching terms, filtering rows and sorting the data. Remember that it is only a viewer: it will never modify your original object.

Notice that in R, the case matters: trying to use `view()` (with a lowercase "V") will yield an error.

> You can also click on the spreadsheet icon in your environment pane to open the viewer.

Finally, to see summary statistics for each of our variables:

```{r}
summary(gapminder)
```

Notice how categorical and numerical variables are handled differently?

## Packages

Packages add functionalities to R and RStudio. There are more than 17000 available.

You can see the list of installed packages in your "Packages" tab, or by using the `library()` function without any argument.

We are going to install and load a new package called "praise". We can do that with the GUI: click the "Install" button in the "Packages" tab (bottom right pane), and search for "praise".

Notice how it runs an `install.packages()` command in the console? You can use that too.

If I try running the command `praise()`, I get an error. That's because, even though the package is installed, I need to **load** it every time I start a new R session. The `library()` function can do that.

```{r}
library(praise) # load the package
praise() # use a function from the package
```

Even though you might need the motivation provided by this function, other packages are more useful for your work.

## Plotting

R already has a number of base plotting functions to visualise data in many different ways. However, many R users will use the package ggplot2 by default because it introduces a logic that makes data visualisation more streamlined, called "Grammar of graphics".

Let's first install ggplot2 on our computer. In the console, execute the following command:

```r
install.packages("ggplot2")
```

We can now load the package, and use the `qplot()` function to visualise the population growth. In our script:

```{r}
library(ggplot2)
qplot(data = gapminder,
      x = year,
      y = pop,
      geom = "point")
```

If we want to add an extra variable, we can add an extra argument:

```{r}
qplot(data = gapminder,
      x = year,
      y = pop,
      colour = continent,
      geom = "point")
```

We can now use the "Export" dropdown menu to save our plot into our "plots" directory in a variety of formats.

## Closing RStudio

You can close RStudio after making sure that you saved your script.

When you create a **project** in RStudio, you create an .Rproj file that gathers information about the state of your project. When you close RStudio, you have the option to save your **workspace** (i.e. the objects in your **environment**) as an .Rdata file. The .Rdata file is used to reload your workspace when you open your project again. Projects also bring back whatever source file (e.g. script) you had open, and your command history. You will find your command history in the "History" tab (upper right panel): all the commands that we used should be in there.

If you have a script that contains all your work, it is a good idea _not_ to save your workspace: it makes it less likely to run into errors because of accumulating objects. The script will allow you to get back to where you left it, by executing all the clearly laid-out steps.

The console, on the other hand, only shows a brand new R **session** when you reopen RStudio. Sessions are not persistent, and a clean one is started when you open your project again, which is why you have to load any extra package your work requires again with the `library()` function.

## What next?

We have a compilation of links for the rest of your R learning: https://gitlab.com/stragu/DSH/blob/master/R/usefullinks.md