---
title: Heatmaps in R for intermediate users
date: 2018-04-20
output:
  html_document: default
  pdf_document: default
  github_document: default
---

> This document is edited as an R markdown file, and regularly exported as a GitHub document.
> See the source code: https://github.com/stragu/CDS/blob/master/R/heatmaps_intermediate.Rmd
> See the published printer-friendly version: 

Thank you for attending this session at the UQ library Centre for Digital Scholarship (CDS).

If you want to review the installation instructions: https://github.com/orchid00/CDS/blob/master/R/Installation.md

Everything we write today will be saved in your R project. Please remember to save it in your H drive or USB if you used the University computers.
Manuals, commands and more information on how to continue your R learning are provided in our community resource in this etherpad: https://etherpad.wikimedia.org/p/cds-heatmaps.

## Keep in mind

* Case sensitive
* No spaces in names
* Use Ctrl + Enter to run a command, and Ctrl + Shift to make your code more readable.

## Open RStudio

* If you are using your own laptop please open RStudio
  * Make sure you have a working internet connection

* On CDS computers:
  * Log in with your UQ username and password
  * Make sure you have a working internet connection
  * Open the ZENworks application
  * Look for the letter R 
  * Double click on RStudio which will install both R and RStudio 
## What are we going to learn?

During this session, you will:

* Learn how to produce a simple heatmap with the base package `heatmap()`;
* Learn about alternatives to produce more complex heatmaps, like `heatmap.2()` and `pheatmap()`;
* Learn how to produce a rudimentary heatmap with the `ggplot2` package.

A **heatmap** is a way of visualizing a table of numbers, where you substitute the numbers with colored cells. It’s useful for finding highs and lows, and see patterns more clearly.

## Disclaimer

We will assume you are an R intermediate user and that you have used RStudio before.

## Material

### RStudio Project

**Exercise 1 - New RStudio Project**

* New project:
  * Click the "New project" menu icon
  * Click "New Directory"
  * Click "New Project" ("New empty project" if you have an older version of RStudio)
  * In Directory name type the name of your project, e.g. "rstudio_intro" (Browse and select a folder where to locate your project, in our case the RProjects folder. If you don't have an RProjects folder, create it.)
  * Click the "Create Project" button
* Create new folders with the following commands:

```{r include=FALSE}
dir.create("scripts")
dir.create("data")
dir.create("plots")
```

**Exercise 2 - setting up**

* create a new R script file called "heatmaps.R" in the "scripts" folder

```{r eval=FALSE, include=FALSE}
file.create("scripts/heatmaps.R")
file.edit("scripts/heatmaps.R")
```

### Method 1: the base `heatmap()` function

*Step 1: import the data*

```{r}
?read.csv
nba <- read.csv(file = "http://datasets.flowingdata.com/ppg2008.csv")
```

*Step 2: explore the data*

```{r eval=FALSE, include=FALSE}
dim(nba)
str(nba)
head(nba)
View(nba)
```

*Step 3: update the dataset*

Assign the Name values to `rownames(nba)`:

```{r}
rownames(nba)
rownames(nba) <- nba$Name
rownames(nba)
```

Delete the column Name:

```{r}
nba$Name <- NULL
head(nba)
```

*Step 4: prepare data, again*

Are you noticing something here? It’s important to note that a lot of visualisations involve gathering and preparing data. Rarely do you get data exactly how you need it, so you should expect to do some data munging before the visuals.

The data was loaded into a data frame, but it has to be a **data matrix** to make your heatmap. The difference between a frame and a matrix is not important for this tutorial. You need to know how to change it.

**Exercise 3 – Convert the nba data into a matrix**

```{r}
class(nba)
nba_matrix <- data.matrix(nba) # convert a DF to a numeric matrix
nba_matrix1 <- matrix(nba) # creates matrix from set of values
nba_matrix2 <- as.matrix(nba) # attempts to turn its argument into a matrix
class(nba_matrix2)
class(nba_matrix1)
class(nba_matrix)
identical(nba_matrix1, nba_matrix) # compare
```

*Step 5: make a heatmap*

We are going to use the `heatmap()` function.

**Exercise 4 – look for the `heatmap()` help**

Read the description of the `scale` argument in particular.

```{r}
?heatmap
```

Create a first heatmap:

```{r}
nba_heatmap <- heatmap(nba_matrix)
```

**Scale** is important: the values should be centered and scaled in either the row direction or the column direction.

```{r}
heatmap(nba_matrix, scale = "column")
```

#### Colours

In the palette function `cm.colours(n)`, n is the number of colors (>= 1) contained in the palette. It can be used in the `col` argument.

In the `margins` arguments, the first value is for column names and the second one for row names.

```{r}
nba_heatmap <- heatmap(nba_matrix,
        col = cm.colors(15),
        scale = "column",
        margins = c(8, 8))
```

#### Remove dendrograms

Dendrograms don't really make sense for this dataset. `Rowv` and `Colv` can be set to `NA` to remove them.

```{r}
nba_heatmap <- heatmap(nba_matrix,
                       col = cm.colors(100),
                       scale = "column",
                       Rowv = NA,
                       Colv = NA)
```

#### Play time!

* Add labels

#### Clean the environment

We can start with a fresh environment, using:

```{r}
rm(list = ls())
```

### Method 2: `gplots::heatmap.2()`

If you don't have the `gplots` package yet, use `install.packages("gplots")`. We will also need the `RColorBrewer`, so use `install.packages("RColorBrewer")` if you don't have it installed already.

```{r}
library(gplots)
library(RColorBrewer)
```

#### Protein data example

*Step 1: data*

Observations for 63 proteins in three control experiments and three experiments where cells are treated with a growth factor.

```{r}
rawdata <- read.csv("https://raw.githubusercontent.com/ab604/heatmap/master/leanne_testdata.csv")
?heatmap.2
```

This `gplots` heatmap function provides a number of extensions to the standard R heatmap function.

*Step 2: explore the data*

```{r}
dim(rawdata)
str(rawdata)
head(rawdata)
View(rawdata)
```

*Step 3: data munging*

```{r}
rawdata$Row_ID <- NULL
rawdata$T.test <- NULL
head(rawdata)
colnames(rawdata) <- c(paste(rep("Control_", 3), 1:3), 
                       paste(rep("Treatment_", 3), 1:3))
head(rawdata)
```

*Step 4: convert the dataframe into a matrix*

```{r}
class(rawdata)
data_matrix <- data.matrix(rawdata)
head(data_matrix)
class(data_matrix)
```

*Step 5: more munging*

```{r}
?scale

```

`scale` is generic function whose default method centers and/or scales the columns of a numeric matrix. We **normalise** the data using `scale()`, **transposing** using `t()` to perform row-wise normalisation of the protein data. This normalises the data across each experiment so we can compare them to each other.
Additionally, we transpose the data back to the original form, for use later on.

```{r}
# Normalise data, transpose for row wise normalisation
data_scaled_t <- scale(t(rawdata))
# check normalisation for each protein by:
  # checking that means are zero
round(colMeans(data_scaled_t), 1)
  # check std devs are 1
apply(data_scaled_t, 2, sd)
# transpose to original form
data_scaled <- t(data_scaled_t)
```

*Step 5: create heatmaps*

First, let's experiment with the base `heatmap()` function.

```{r}
heatmap(data_scaled)
```

```{r}
heatmap(data_scaled,
        col = heat.colors(8), 
        margins = c(8,12)) # , reveal col labels, reduce width
```

Let's create a new palette with 25 increments:

```{r}
my_palette <- colorRampPalette(c("blue",
                                 "white",
                                 "red"))(n = 25)
```

Now, let's use `heatmap.2()`:

```{r}
dev.off()
par(cex.main = 0.75)                    # Shrink title fonts on plot
heatmap.2(data_scaled,                  # normalised data
          density.info = "histogram",   # Plot histogram of data and colour key
          trace = "none",               # Turn off trace lines from heatmap
          col = my_palette,             # Use my colour scheme
          cexRow = 0.5, cexCol = 0.75)  # Amend row and column label fonts
```

Fix a few things and add a few extras:

```{r}
heatmap.2(data_scaled,
          density.info = "histogram",
          Colv = NA,                    # only row dendrogram
          trace = "none",
          col = my_palette,
          main = "Difference between proteins",
          margins = c(8, 5),            # more space from border
          offsetRow = 1,                # space between plot and labels
          keysize = 2,                  # make key and histogram fit
          cexRow = 0.5, cexCol = 0.75)
```

Clean up the environment with:

```{r}
rm(list = ls())
```

#### Genechip data example

The `BiocInstaller` package allows to use the `biocLite()` function to install extra packages, including the `affy` package which we need in this example (it contains the data).
Install BiocInstaller with `source("https://bioconductor.org/biocLite.R")`, and then `affy` with the command `biocLite("affy")`

```{r eval=FALSE, include=FALSE}
library(affy)
data(SpikeIn)
pms <- SpikeIn@pm
str(pms)
class(pms)
# just the data, scaled across rows
par(cex.main = 0.75)  
labs <- colnames(pms)
heatmap.2(pms, 
          col = rev(heat.colors(16)), 
          main = "SpikeIn@pm",
          xlab = "Relative Concentration", 
          ylab="Probeset",
          trace = "none",
          labCol = "",
          add.expr = text( x = seq_along(labs), y = -2, srt = 45,
                           labels = labs, xpd = TRUE),
          scale="row")
```

Clean up your environment with:

```{r}
rm(list = ls())
```

### Method 3: `pheatmap()`

If you don't have it already, install `pheatmap` with `install.packages("pheatmap")`.
Load the packages with:

```{r}
library(grid)     # this is a base package  
library(pheatmap) # stands for "pretty heatmaps"
```

How does `pheatmap` differ?

```{r}
?pheatmap
```

> A function to draw clustered heatmaps where one has better control over some graphical parameters such as cell size, etc.

Create your data matrix:

```{r}
set.seed(100)
d <- matrix(rnorm(25), 5, 5)
colnames(d) <- paste("Treatment", 1:5, sep = "")
rownames(d) <- paste("Gene", 1:5, sep = "")
```

Try it out:
```{r}
pheatmap(d, 
         main = "Pretty heatmap",
         cellwidth =  50,
         cellheight = 30,
         fontsize = 12,
         filename = "plots/heatmap.pdf")
```

Clean up your environment with:

```{r}
rm(list = ls())
```

### Method 4: a dataframe in `ggplot2`

Load the necessary libraries:

```{r}
library(reshape2)
library(ggplot2)
```

Labels the rows and columns:

```{r}
name_genes <- paste(rep("GEN", 20), LETTERS[1:20], sep="_") # rows
name_patients <- paste(rep("PATIENT", 20), seq(1,20,1), sep="_")  # columns
```

Generation of dataframe:

```{r}
value_expression <- data.frame(genes = name_genes, 
                               matrix(rnorm(400, 2, 1.8),
                                      nrow = 20,
                                      ncol = 20))
names(value_expression)[2:21] <- name_patients
```

 
Melt dataframe:
```{r}
df_heatmap <- melt(value_expression, id.vars = "genes")
names(df_heatmap)[2:3] <- c("patient", "expression_level")
head(df_heatmap)
```

```{r}
ggplot(df_heatmap, aes(patient, genes)) +
  geom_tile(aes(fill = expression_level), color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  ylab("List of genes") +
  xlab("List of patients") +
  theme(legend.title = element_text(size = 10),
        legend.text = element_text(size = 12),
        plot.title = element_text(size=16),
        axis.title=element_text(size=14,face="bold"),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(fill = "Expression level")
```

Clean up your environment with:

```{r}
rm(list = ls())
```

### Method 5: `ComplexHeatmap::Heatmap`

URL: https://www.bioconductor.org/packages/devel/bioc/vignettes/ComplexHeatmap/inst/doc/s2.single_heatmap.html

*Step 1*

```{r}
# source("https://bioconductor.org/biocLite.R")
# biocLite("ComplexHeatmap") 
library(ComplexHeatmap)
# install.packages("circlize")
library(circlize)
```

Create a data matrix:

```{r}
set.seed(123)
mat <- cbind(rbind(matrix(rnorm(16, -1), 4), matrix(rnorm(32, 1), 8)),
            rbind(matrix(rnorm(24, 1), 4), matrix(rnorm(48, -1), 8)))
```

*Step 2: munging*

Permute the rows and columns
```{r}
mat <- mat[sample(nrow(mat), nrow(mat)), sample(ncol(mat), ncol(mat))]
rownames(mat) <- paste0("R", 1:12)
colnames(mat) <- paste0("C", 1:10)
```

*Step 3: make a heatmap*

```{r}
?Heatmap
Heatmap(mat)
```

```{r}
Heatmap(mat, 
        col = colorRamp2(c(-3, 0, 3), c("brown", "white", "yellow")), 
        cluster_rows = FALSE, 
        cluster_columns = FALSE,
        heatmap_legend_param = list(title = "Values"))
```

```{r}
Heatmap(mat, 
        col = colorRamp2(c(-3, 0, 3), c("brown", "white", "yellow")), 
        heatmap_legend_param = list(title = "Values"))
```

```{r}
mat_with_na <- mat
mat_with_na[sample(c(TRUE, FALSE), nrow(mat)*ncol(mat), replace = TRUE, prob = c(1, 9))] <- NA
Heatmap(mat_with_na, 
        col = topo.colors(100),
        na_col = "orange", 
        clustering_distance_rows = "pearson",
        heatmap_legend_param = list(title = "Values"))
```


```{r}
Heatmap(mat, 
        name = "values", 
        row_names_side = "left", 
        row_dend_side = "right", 
        column_names_side = "top", 
        column_dend_side = "bottom")
```

```{r}
Heatmap(mat, 
        name = "values", 
        row_names_side = "left", 
        row_dend_side = "right", 
        column_names_side = "top", 
        column_dend_side = "bottom",
        km = 2)
```

```{r}
h1 <- Heatmap(mat, 
        name = "values", 
        #col = topo.colors(50),
        color_space = "sRGB",
        row_dend_width = unit(2, "cm"),
        column_dend_height = unit(2, "cm"),
        row_dend_reorder = TRUE,
        column_dend_reorder = TRUE,
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 7),
        column_names_max_height = unit(2, "cm"),
        row_names_max_width = unit(9, "cm"),
        column_title = "This is a complex heatmap")
draw(h1, heatmap_legend_side = "left")
```


Clean my environment with:

```{r}
rm(list = ls())
```

## Close Rproject

- File
- close project
- then close RStudio

## Important links
* RStudio Cheatsheet https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf
* RStudio online learning https://www.rstudio.com/online-learning/
* Basic and advanced manuals https://cran.r-project.org/manuals.html
* Ask about any function or package http://www.rdocumentation.org/
* If you are looking how-to's or how to fix an error http://stackoverflow.com/questions/tagged/r
* Lynda.com R training and tutorials https://www.lynda.com/R-training-tutorials/1570-0.html remember to sign in with your organisational portal,https://web.library.uq.edu.au/library-services/training/lyndacom-online-courses
* R colours http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
* Book: Hadley Wickham. ggplot2 Elegant Graphics for Data Analysis Second  Edition. 2016 
https://link-springer-com.ezproxy.library.uq.edu.au/content/pdf/10.1007%2F978-3-319-24277-4.pdf
* Examples of heatmaps: https://flowingdata.com/2010/01/21/how-to-make-a-heatmap-a-quick-and-easy-solution/
 https://rpubs.com/ab604/98032  https://stackoverflow.com/questions/15505607/diagonal-labels-orientation-on-x-axis-in-heatmaps  https://www.bioconductor.org/packages/devel/bioc/vignettes/ComplexHeatmap/inst/doc/s2.single_heatmap.html

* If you need an R and/or RStudio workshop/session, please contact Centre for Digital Scholarship staff to organise one for you. https://web.library.uq.edu.au/locations-hours/centre-digital-scholarship
* If you have further questions, please contact me p.martinez at uq.edu.au 
